\ifdefined\included
\else
\documentclass[a4paper,11pt,twoside]{StyleThese}
\include{formatAndDefs}
\sloppy
\begin{document}
\fi


\chapter*{Conclusion and Perspectives}
\addstarredchapter{Conclusion and Perspectives} %Sinon cela n'apparait pas dans la table des mati?res
\setcounter{secnumdepth}{-1}
\chaptermark{Conclusion and Perspectives}

\cleardoublepage

In the frame of the \koroibot\ project, researchers integrated and developed new controllers to improve humanoid robot locomotion.
Collaborations with the \koroibot\ partner universities and the Airbus industry were fruitful.
In this context, the contributions of this thesis can be classified in two parts.
On one hand, there are theoretical contributions concerning the locomotion of humanoid robot.
On the other hand, technical contributions were done in industrial and academic context.
In the next section we will summarize the work that has been explained in this thesis.
And in a second part we will present the further improvements that can be made to increase the performance of the existing algorithms.

\input{conclusion_contribution.tex}
\input{conclusion_perspective.tex}

This thesis shows that humanoid robot flat ground walking is a topic nearly solved.
Application using high level controller can already be demonstrated.
The extension of this work, the multicontact locomotion, is a very important topic nowadays.
It captures the idea of a generic algorithm that can be used for flat ground locomotion or parkour motion execution.
There are plenty of room for improvement in this thesis work and interesting perspectives concerning the robot locomotion performance and versatility.
Other European project like the \koroibot\ project should arise to see these improvements becoming a reality.

\setcounter{secnumdepth}{+3}

\ifdefined\included
\else
{
\bibliographystyle{StyleThese}
\newgeometry{left=1in,right=1in,top=1in,bottom=1in,
  includefoot,includehead,headheight=13.6pt}
\bibliography{16-thesis-mnaveau-short}
\restoregeometry
}
\end{document}
\fi
